\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-10-06 doc/paper-jss/sdmTMB-paper.Rnw
+ paragraph on applications to other fields + more refs in bib + minor word changes (mostly intro, clarity)
* Noticed there's a mix of \sdmTMB\ and \pkg{sdmTMB} throughout. Change throughout for consistency?

\Abstract{
Geostatistical data are common across scientific fields and available in increasingly large datasets. 
However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging to implement for many users. 
Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. 
SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}. 
We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation). 
Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. 
Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \pkg{sdmTMB} can be an order of magnitude faster than \proglang{R}-\pkg{INLA}. 
We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
}

@@

Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects are a class of models appropriate for these data \citep{rue2005gmrf, diggle2007, cressie2011}. 
Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time. 
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by some covariance function such as the exponential or Matérn \citep{cressie1993, diggle2007, chiles1999}.
% TODO: decide on using GRF throughout

Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data. 
Many solutions have been proposed, such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011}, and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}. 
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian \citep{gelfand2017} or maximum likelihood methods. 
This can greatly improve computational efficiency \citep[e.g.,][]{heaton2019}. 
The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \pkg{TMB} \citep[Template Model Builder,][]{kristensen2016} that relies on \proglang{R}-\pkg{INLA} to create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.

@@

Software packages for specifying statistical models that can include the SPDE, such as \proglang{R}-\pkg{INLA} and \pkg{TMB}, are flexible and powerful but are challenging to use for many applied researchers. 
For example, \pkg{TMB} requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code. 
Packages such as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE or spatiotemporal functionality. 
Packages such as \pkg{VAST} \citep{thorson2019} and \pkg{inlabru} \citep{bachl2019} are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, or lack some functionality.

@@

Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \INLA\ to set up SPDE matrices. 
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models. 
Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package. 
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.

This paper describes the underlying statistical models (Section~\ref{model-description}), explains how \pkg{sdmTMB} is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \pkg{sdmTMB} to other packages (Section~\ref{package-comparisons}) and concludes with a discussion of TODO  (Section~\ref{discussion}).

@@

The design goal of \sdmTMB\ was to develop a flexible, modular, and familiar inteface to fast maximum likelihood inference (or full Bayesian inference) with the SPDE approach to spatial and spatiotemporal GLMMs with random fields.
The package gathers functionality not found combined in other packages that is particularly useful to species distribution modelling, but is applicable beyond ecology to any field that encounters geostatistical data that is continuously referenced in space and (optionally) discretely indexed by time.

@@

\pkg{sdmTMB} models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \pkg{TMB} \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the spatial range and variance parameters.
These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan}, a \texttt{bayesian} logical argument is available to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

TODO add paragraph here on applications to other fields, beyond ecology and fisheries? Specifically thinkgin about:
- spatial econometrics (e.g. Elhorst, 2012; Lee \& Yu, 2010)
- disease and medicine (e.g. Wakefield work)
- agriculture (https://link.springer.com/article/10.1007/s00122-019-03424-y)

>>>>

\Abstract{
Geostatistical data are common across scientific fields and the dimensions of these data continue to become larger in space and time.
However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging for many users to implement.
Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation).
Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation.
Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \pkg{sdmTMB} can be an order of magnitude faster than \proglang{R}-\pkg{INLA}.
We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
}

@@

Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects constitute an appropriate class of models for such data \citep{rue2005gmrf, diggle2007, cressie2011}.
Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables that cause observations to be correlated in space or both space and time.
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by specific covariance functions such as the exponential or Matérn \citep{cressie1993, diggle2007, chiles1999}.
% TODO: decide on using GRF throughout

Such models quickly become computationally challenging due to the need to invert large matrices to account for covariation among data.
Many solutions have been proposed, such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011}, and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}.
These approaches aim to minimize the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian \citep{gelfand2017} or maximum likelihood methods.
This can greatly improve computational efficiency \citep[e.g.,][]{heaton2019}.
The SPDE approach has been widely adopted, especially via the \proglang{R}-\pkg{INLA} package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \pkg{TMB} \citep[Template Model Builder,][]{kristensen2016} that relies on \proglang{R}-\pkg{INLA} to create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.

@@

Software packages designed for specifying statistical models which incorporate the SPDE, such as \proglang{R}-\pkg{INLA} and \pkg{TMB}, are flexible and powerful but can be challenging for many applied researchers.
For example, \pkg{TMB} requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
While packages such as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---they lack built-in SPDE or spatiotemporal functionality.
Packages such as \pkg{VAST} \citep{thorson2019} and \pkg{inlabru} \citep{bachl2019} are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, or lack some functionality.

@@

Here, we introduce the \R\ package \pkg{sdmTMB}, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \INLA\ to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
Many features of \pkg{sdmTMB} may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: \pkg{sdmTMB}), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.

This paper describes the underlying statistical models (Section~\ref{model-description}), explains how \pkg{sdmTMB} is designed and summarizes its software functionality (Section~\ref{sec:software}), illustrates its use through three case studies (Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}), compares \pkg{sdmTMB} to other packages (Section~\ref{package-comparisons}) and concludes with a discussion of links to other packages and future development directions (Section~\ref{discussion}).

@@

The design goal of \sdmTMB\ was to develop a flexible, modular, and intuitive inteface to fast maximum likelihood inference (or full Bayesian inference) with the SPDE approach to spatial and spatiotemporal GLMMs with random fields.
The package gathers functionality not found combined in other packages that is particularly useful to species distribution modelling, but is applicable beyond ecology to any field that encounters geostatistical data that is continuously referenced in space and (optionally) discretely indexed by time.

@@

\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \pkg{TMB} \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the spatial range and variance parameters.
These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan}, a \texttt{bayesian} logical argument is available to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

While our focus is primarily in applications to biological data, the SPDE approach and functionality of \sdmTMB\ has applications in many other fields. Examples include spatial models of disease spread \citep{moraga2021}, spatial econometric models of quantities such as housing prices \citep{bivand2014}, analyzing medical imaging data such as MRI scans \citep{naseri2022}, and geophysical models of seismic waves following earthquakes \citep{zhang2015}. These examples underscore the versatile nature of the SPDE approach and the potential uses of \sdmTMB\ across various scientific and industrial sectors.
